---
title: "R Basics"
tutorial:
  id: "yousaydata.rbasics"
  version: 2.0.0
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: simplex
    includes:
      before_body: "./www/branding.html"
runtime: shiny_prerendered
description: >
  Learn some basic features of R.
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
gradethis::gradethis_setup()
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(collapse = TRUE)
```


## 1. The Very Basics

R is like a fast calculator

```{r echo=T}
232047294879347 * 242424 / 12313 + 2

```

Here R only does the calculation for you. It does not actually save anything to memory. In R `<-` is used to assign a value to a variable. You can also use `=`, but this is considered not good style. If you are concerned that you have to type an extra sign use the [keyboard shortcut](https://support.rstudio.com/hc/en-us/articles/200711853-Keyboard-Shortcuts) `option/alt + -`, which will produce `<-`. 

```{r echo=T}
myvar <- 232047294879347 * 242424 / 12313 + 2
```

As you can see, nothing is being printed to the console, but you will be able to see in RStudio that a variable has been added to your `Globale Environment`:

```{r}
values <- tibble(`Values` = "myvar", ` ` = 232047294879347 * 242424 / 12313 + 2)
knitr::kable(values)
```

If you want to save a variable to memory *and* print it out you have to wrap the assignment in parentheses:

```{r echo=T}
(myvar <- 232047294879347 * 242424 / 12313 + 2)
```

Try it for yourself Assign `2` to variable that you call `a`:

```{r ex1_1, exercise = TRUE}

```

```{r ex1_1-solution}
a <- 2
```
 
```{r ex1_1-code-check}
grade_code()
```

Once you have assigned a variable you can use it for updating the variable itself. What would be the result of this (replace the `_` with the solution)?

```{r ex1_2, exercise = TRUE}
a <- 2
a <- a * a * a
a == _
```

```{r ex1_2-solution}
a <- 2
a <- a * a * a
a == 8
```
 
```{r ex1_2-code-check}
grade_code()
```

Aside from Base R functions, you can include third party functions, objects, and data via libraries:

```{r echo=T}
# install.packages("tidyverse")
# This without the leading # would be the way to install a library.
# A line with a leading # is ignored by R.
# I have done so, because I do not want to install a package, already installed on my computer.
# You only have to use the install.packages() function once on each new system
# or after upgrading your R version.

# You can also use those #-lines to comment your code
# if you are using RStudio have a look what happens when you
# start a line with # and finish the line with ####
# This feature of RStudio helps you to organise your code.
library(tidyverse)
```

Which gives you access to more functionality for your projects:

```{r echo=T}
mpg %>% ggplot(aes(displ,hwy)) + geom_point()
```

You do not need to understand the above code yet, just understand that we are using an external library. The `tidyverse` package is a library that contains a lot of useful features and it is the package, I use in almost every R program.

That being said, you can also generate your own functions:

```{r echo=T}
myFunction <- function(x) {
  result <- x + 2
  return(result)
}
myFunction(15)
```

## 2. Data Types

The datatype of each variable is inferred:  
Logical is either `TRUE` or `FALSE`.
```{r echo=T}
variable <- FALSE 
class(variable)
```

Numeric is a number that can include fractual parts, i.e. 2.34.

```{r echo=T}
variable <- 2.34 
class(variable)
```

Integer is a whole number.

```{r echo=T}
variable <- 2L 
variable
class(variable)
```

### Sidetrack

Integers are cheaper than Numerics:

```{r echo=T}
object.size(c(1L,2L,3L,4L,5L)) < object.size(c(1,2,3,4,5))
```

Unless they are empty:

```{r echo=T}
object.size(numeric()) == object.size(integer())
```

Then they are both 48bytes:

```{r echo=T}
object.size(numeric())
```

### Back on track

You can also express complex numbers.

```{r echo=T}
variable <- 3 + 2i
variable
class(variable)
```

And plot them:

```{r echo=T}
plot(variable)
```

Strings are called characters in R.

```{r echo=T}
variable <- "hello world"
variable
class(variable)
```

The last data type is `raw`. This refers to a byte stream.

```{r echo=T}
variable <- charToRaw(variable)
variable
class(variable)
```

Raw objects are small. Empty raw objects are also 48bytes though:

```{r echo=T}
object.size(raw())
```

The types you will use most are characters, numeric, and logical.

## 3. Data Structures

### Vectors

Technically everything that you have produced so far are vectors. You build a vector using the `c` function:

```{r echo=T}
(myVector <- c(1,2,4,5))
```

And you can extract elements by index using `[]`:
```{r echo=T}
myVector[3]
myVector[2:3]
myVector[1]
```

Vectors can be combined with other vectors using the `c()` function. Vectors can only contain *one type* of data. They will throw an error or normalise if you try to force two different types together.

```{r echo=T}
VectorA <- 2
VectorB <- c(1L,2L)
class(VectorB)

c(VectorA,VectorB)
class(c(VectorA,VectorB))
```

### Lists

A list is an R-object which can contain many different types, such as vectors, functions and even another list inside it.

```{r echo=T}
list(VectorA, VectorB, list(VectorA, VectorB), mean)
```

Lists can be named:

```{r echo=T}
mylist <- list(a = VectorA, b = VectorB, c = list(VectorA, VectorB), d = mean)
```

And you can extract elements either by their names using the `$` operator or by index using double square brackets `[[]]`.

```{r echo=T}
mylist$b
mylist[[1]]
```

N.B. `[[]]` also work for vectors if you do not want to remember the difference.

### Matrices

A matrix is a two-dimensional rectangular data set of ONE data type. It can be created using a vector input to the matrix function. This is how you create a matrix:

```{r echo=T}
M <- matrix( c('a','a','b','c','b','a'), nrow = 2, ncol = 3, byrow = TRUE)
M
```

If you want to know who the `matrix()` function works, you can ask R:

```{r echo=T, eval=F}
?matrix
```

Let's have a look at the result, if we use the same input vector, but a different method of filling the matrix:

```{r echo=T}
M <- matrix(c('a','a','b','c','b','a'), nrow = 2, ncol = 3, byrow = FALSE)
M
```

Do you see how it has changed? 

You can extract rows from a matrix:

```{r echo=T}
M[1,]
```

You can extract columns from a matrix:

```{r echo=T}
M[,1]
```

Or individual cells

You can extract rows from a matrix:

```{r echo=T}
M[1,2]
```

### Factors

A factor stores the vector and all its unique values. Let's start with creating a vector.

```{r echo=T}
# Create a vector.
apples <- c('iOS','iPhone','MBP','iPhone','MBP')
```

The vector above has 5 elements total, but only 3 unique elements. If nothing is specified a factor will display all the options a value could have in the vector.

```{r echo=T}
# Create a vector.
factor(apples)
```

Instead of infering it from a given vector, we could also supply a vector with all potential values and even supply an order:

```{r echo=T}
# Create a factor object.
factor_apples <- factor(apples, levels = c("iOS", "iPhone", "MBP", "Mac"), ordered = T)
```

Compare the outputs:

```{r echo=T}
factor(apples)
factor_apples
nlevels(factor_apples)
```

We can extract individual elements using `[]`. Note that R will tell us all the levels as well. If we want to only extract the value we have to turn it into a string:

```{r echo=T}
factor_apples[4]
as.character(factor_apples[4])
```

### Data Frames

And finally, Data Frames are mixed data presented in tabular form. 

```{r echo=T}
# Create the data frame.
data_types <- data.frame(
  types = c("logical", "numerical","integer", "complex", "character", "raw"), 
  usage_probability = c(1, 1, 1, 0.1, 1, 0.5), 
  got_it = c(rep("yes", 5), "no"),
  answer = rep(42L, 6)
)
```

While we can print it in a pretty way (which this renderer does by default):

```{r echo=T}
data_types
```

The `normal` print in the console is less useful:
```{r echo=T}
knitr::normal_print(data_types)
```


The `tidyverse` has also a modernised version of the dataframe called tibble, which behaves nicer:

```{r echo=T}
library(tidyverse)
data_types_tibble <- as_tibble(data_types)

data_types_tibble
```

and in the console:

```{r echo=T}
knitr::normal_print(data_types_tibble)
```

You might already see the advantages of the tibble.

From a computer science perspective dataframes are a special type of list. They are a list where each element has exactly the same length as all other elements (that is what makes them rectangular). Because of this, the `$` operator that we know from named lists works for them too and because they are rectangular we can also use indices like with matrices:

```{r echo=T}
data_types$usage_probability

data_types[,1:2]
```


## 4. Other Data Structures

### Strings and Arrays

Strings in R are called characters and they behave differently than in other programming languages:

```{r echo=T}
newString <- "test"
length(newString)
nchar(newString)
```

One dimensional arrays of simple types in R are best expressed as vectors. And so, if another developer talks about "arrays", the R programmer should translate that to "vectors". There is actually a data structure called "array" in R, but this is mainly reserved for multi-dimensionional arrays:

```{r echo=T}
intArray <- c(7L,8L,9L,10L)
length(intArray)
intArray 

## sidetrack, you do not need to remember this
intRealArray <- array(c(7L,8L,9L,10L),dim = c(2,2))
intRealArray
```

### Hash table

R does not have hash tables. Instead, R uses something like it that is called environment or it uses lists. That said ther is a `hashmap` library.

### Linked Lists

Linked list are directly implemented in R using list:

```{r echo=T}
lst <- list() # creates an empty (length zero) list
lst
```

```{r echo=T}
lst[[1]] <- 1 # automagically extends the lst
lst[[2]] <- 2 # ditto
lst
```

You can also create an empty list of  a set length:

```{r echo=T}
lst <- vector("list", 5)
lst
```

Then populate it:

```{r echo=T}
lst <- list(1, 2, 3, 4, 5)
lst
```


Or delete individual elements:

```{r echo=T}
lst <- lst[-2]
lst
```

### Trees

In R trees are just a list of lists. As an aside, R sometimes is not very memory efficient (if you do not watch out) and it likes copying (a lot). And so, if you want to store the data as a more complex graph, you need to use a graph db.

Here is an example of a simple tree:

```{r echo=T}
tree <- list(
  node1 = list(
    1, 2), 
  node2 = list(
    node1 = 3, 
    node2 = list(
      node1 = 4, 
      node2 = 5)))
```

You can extract it using the `[[]]` operator

```{r echo=T}
# left child of right child of right child of parent: 4
tree[[2]][[2]][[1]]
```

Or maybe more convenient the names:

```{r echo=T}
tree$node2$node2$node1
```

That's all. Feel free to go over this page again at a later stage. Those are some solid mechanics for more complex problems.