---
title: "Intro to Visualisations in R"
tutorial:
  id: "yousaydata.introvis"
  version: 2.0.0
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: simplex
    includes:
      before_body: "./www/branding.html"
runtime: shiny_prerendered
description: >
  Learn some basic concepts of ggplot2.
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
gradethis::gradethis_setup()
knitr::opts_chunk$set(echo = FALSE)
```


## 1.1 The library we use

One of the most central libraries in modern R is `tidyverse`. We are going to use it throughout. `tidyverse` is essentially a collection of many different libraries for working with rectangular data.

```{r echo=TRUE}
library(tidyverse)
```

In what follows let's try to answer whether cars with big engines use more fuel than cars with small engines? 

For this we are going to use the `mpg` data set that is included in `tidyverse`. We can just see what it is by calling it: 

```{r lookup, exercise=TRUE}
mpg
```

Since it is part of an official package it is also documented and we can use the `?` command to see the documentation (that is nicer in RStudio)":

```{r lookup2, exercise=TRUE}
?mpg
```

If we want to plot it, we can do it using `ggplot`:

```{r firstplot, exercise=TRUE}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy))
```

So we see that `ggplot` has the following simple structure:

```{r echo=T}
# ggplot(data = <DATA>) +
# <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))
```

## 1.2 Exercises

1. Run `ggplot(data = mpg)`. What do you see?

```{r ex121, exercise=TRUE}

```

2. How many rows are in mpg? How many columns?

```{r ex122, exercise=TRUE}

```

3. What does the drv variable describe? Read the help for ?mpg to find out.

```{r ex123, exercise=TRUE}

```

4. Make a scatterplot of hwy vs cyl (replace the `<...>` in the code).

```{r ex124, exercise=TRUE}
ggplot(data = <DATA>) +
<GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))
```

5. What happens if you make a scatterplot of class vs drv? Why is the plot not useful?

```{r ex125, exercise=TRUE}
ggplot(data = <DATA>) +
<GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))
```


## 2.1 Aesthetics 

Let's have a look a how we can visuallly mark some outliers:

```{r}
outliers <- (mpg$hwy > 20) & (mpg$displ > 5)
ggplot(mpg, aes(x = displ, y = hwy, colour = outliers)) +
  geom_point()
```

We can see that aside from x and y, there are other aesthetics of the visualisation that I can change. For instance the colour.

```{r colmark, exercise=T}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = class))
```

Of course there are more aesthetics and options.

Like size:
```{r sizemark, exercise=T}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = class))
```

Like the alpha value of a colour:
```{r alphamark, exercise=T}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, alpha = class))
```

Or shape:
```{r shapemark, exercise=T}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape = class))
```

If I want to make a general change, I am doing this outside of the `aes` function:
```{r allmark, exercise=T}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), size = 5)
```

## 2.2. Exercises

1. What’s gone wrong with this code? Why are the points not blue? Can you correct it?

```{r ex221, exercise=T}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = "blue"))
```

2. Which variables in mpg are categorical? Which variables are continuous? (Hint: type ?mpg to read the documentation for the dataset). How can you see this information when you run mpg?

```{r ex222, exercise=T}

```

3. Map a continuous variable to color, size, and shape. How do these aesthetics behave differently for categorical vs. continuous variables?

```{r ex223, exercise=T}

```

4. What happens if you map the same variable to multiple aesthetics?

```{r ex224, exercise=T}

```

5. What does the stroke aesthetic do? What shapes does it work with? (Hint: use ?geom_point)

```{r ex225, exercise=T}

```

6. What happens if you map an aesthetic to something other than a variable name, like aes(colour = displ < 5)? Note, you’ll also need to specify x and y.

```{r ex226, exercise=T}

```

## 2.3. Facets

We can also use facets for our visualisations. Either we wrap it by one variable with `facet_wrap`:

```{r echo=T}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2)
```

Or produce a two-dimensional grid with `facet_grid`:

```{r echo=T}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(drv ~ cyl)
```

Feel free to try it yourself:

```{r ptt23, exercise = T}

```

## 3.1 GEOMs

I am sure you do not want to just produce scatter plots all the time. `ggplot` has you covered for all kinds of plots:

What's the difference here?

```{r echo=T}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))

ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy))
```

With other types of plots, `ggplot` gives you access to another set of aesthetics. For instance, the `linetype` aesthetic:

```{r echo=T}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv))
```

In addition to switching `geom_`-functions you can also visually group the data or modify the legend. Try any of those:

```{r ptt31, exercise = T}
# ggplot(data = mpg) +
#  geom_smooth(mapping = aes(x = displ, y = hwy))
#
# ggplot(data = mpg) +
#  geom_smooth(mapping = aes(x = displ, y = hwy, group = drv))

# ggplot(data = mpg) +
#   geom_smooth(
#     mapping = aes(x = displ, y = hwy, color = drv),
#     show.legend = FALSE
#  )
```

It is also possible to use multiple different `geom`-functions. You essentially stack the visualisations on top of each other. `ggplot` works in layers!

```{r, echo=T}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  geom_smooth(mapping = aes(x = displ, y = hwy))
```

Try to avoid code duplication, though. The `geom`-functions can inhere from their parent `ggplot`-function. Checkt it out yourself: 

```{r ptt31a, exercise=T}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()
```

If you do this, you can always overwrite inhereted settings for individual layers. What do you think this will produce? 

```{r ptt31b, exercise=T}
ggplot(mpg, aes(x = displ, y = hwy, col = drv)) + 
  geom_point(aes(color = class), show.legend = F) +
  geom_smooth()
```

(Needless to say, it is not a very useful plot)

In any case, you can even switch the data you are using:

```{r ptt31c, exercise=T}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth(data = filter(mpg, class == "subcompact"), se = FALSE)
```

## 3.2. Exercises

1. What geom would you use to draw a line chart? A boxplot? A histogram? An area chart?

```{r ex321, exercise=T}

```

2. Run this code in your head and predict what the output will look like. Then, run the code in R and check your predictions.

```
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) + 
  geom_point() + 
  geom_smooth(aes(linetype = drv), se = FALSE)
``` 

```{r ex322, exercise=T}

```

3. What does show.legend = FALSE mean? What happens if you remove it? 

```{r ex323, exercise=T}
ggplot(data = mpg) +
  geom_smooth(
    mapping = aes(x = displ, y = hwy, color = drv),
    show.legend = FALSE
  )
```

4. What does the se argument to geom_smooth() do?

```{r ex324, exercise=T}

```